# Assignment 2

## Load all libraries
```{r, warning=FALSE}
library("org.Mm.eg.db")
library(magrittr)
library(readr)
library(dplyr)
library(tidyr)
library(biomaRt)
library(biomartr)
library(ggfortify)
library(DESeq2)
library(ggplot2)
library(tibble)
library(EnhancedVolcano)
library(topGO)
library(clusterProfiler)
library(gprofiler2)
library(ComplexHeatmap)
library(umap)
```

## Read in data
Data will be in a folder called "SROP119064".
```{r warning=FALSE}
# Read in metadata
metadata <- read_tsv("SRP119064/metadata_SRP119064.tsv")

# Read in data
expression_df <- read_tsv("SRP119064/SRP119064.tsv")%>%
  # Convert Gene ID column to rowname
  column_to_rownames("Gene")

# Align data and metadata
# Select from expression_df the rows with metadata's refinebio_accession_code
expression_df <- expression_df %>%
  dplyr::select(metadata$refinebio_accession_code)

# Check that they are in the same order
all.equal(colnames(expression_df), metadata$refinebio_accession_code)

expression_df <- expression_df %>%
  rownames_to_column("Gene")

mart <- useMart('ensembl', dataset = 'mmusculus_gene_ensembl')
```

## Get gene names
```{r}
mapping_ensembl_to_symbol <- getBM(
  attributes = c('ensembl_gene_id', 'mgi_symbol', 'go_id'),
  values = expression_df$Gene,
  mart = mart)

collapsed_mapping_df <- mapping_ensembl_to_symbol %>%
  # Group by Ensembl IDs
  group_by(ensembl_gene_id) %>%
  # Collapse the mgi symbols `mapping_df` into one column named `all_symbols`
  summarize(all_symbols = paste(unique(mgi_symbol), collapse = ";"),
            go_ids = paste(unique(go_id), collapse = ";"))

final_mapping_df <- collapsed_mapping_df %>%
  # Add the rest of the expression data
  dplyr::right_join(expression_df, by = c("ensembl_gene_id" = "Gene"))
```

## Write data to output file
```{r}
# Write data frame to output file
write_tsv(final_mapping_df, file.path(
  "results/SRP040561_GeneNames.tsv"
))
```

## Create density plot
```{r}
logged_mapping <- final_mapping_df %>% 
  column_to_rownames("ensembl_gene_id") %>%
  dplyr::select_if(is.numeric) %>%
  log()

per_gene_median_expressions <- logged_mapping %>%
  rowwise() %>%
  mutate(per_gene_medians = median(c_across(where(is.numeric)), na.rm=TRUE)) %>%
  dplyr::select(per_gene_medians)

ggplot(per_gene_median_expressions, aes(x=per_gene_medians)) + 
  geom_density()
```

## Create PCA plot
```{r}
final_mapping_df <- final_mapping_df %>%
  column_to_rownames("ensembl_gene_id") %>%
  dplyr::select(metadata$refinebio_accession_code)

# Check that they are in the same order
all.equal(colnames(final_mapping_df), metadata$refinebio_accession_code)

gene_matrix <- round(final_mapping_df)

metadata <- metadata %>%
  # Get trem2knockout status
  dplyr::mutate(knockout_status = dplyr::case_when(
    stringr::str_detect(refinebio_subject, "wt") ~ "wt",
    stringr::str_detect(refinebio_subject, "trem2ko") ~ "trem2ko"
  )) %>%
  # Get time passed (4 months or 8 months)
  dplyr::mutate(time_passed = dplyr::case_when(
    stringr::str_detect(refinebio_subject, "4m") ~ "4m",
    stringr::str_detect(refinebio_subject, "8m") ~ "8m"
  ))
metadata <- metadata %>%
  dplyr::mutate(
    # Here we define the values our factor variable can have and their order.
    knockout_status = factor(knockout_status, levels = c("wt", "trem2ko"))
  ) %>%
  dplyr::mutate(
    refinebio_specimen_part = factor(refinebio_specimen_part, levels = c("cortex", "hippocampus"))
  ) %>%
  dplyr::mutate(
    time_passed = factor(time_passed, levels = c("4m", "8m"))
  )

ddset <- DESeqDataSetFromMatrix(gene_matrix, metadata, ~refinebio_specimen_part)
deseq <- DESeq(ddset)

plotPCA(DESeqTransform(deseq), intgroup = "refinebio_specimen_part")
```

## UMAP Plot
```{r}
gene_norm <- vst(ddset)
normalized_counts <- assay(gene_norm) %>%
  t() # We need to transpose this data so each row is a sample
umap_results <- umap::umap(normalized_counts)
umap_plot_df <- data.frame(umap_results$layout) %>%
  # Turn sample IDs stored as row names into a column
  tibble::rownames_to_column("refinebio_accession_code") %>%
  # Add the metadata into this data frame; match by sample IDs
  dplyr::inner_join(metadata, by = "refinebio_accession_code")
umap_plot_df
ggplot(
  umap_plot_df,
  aes(
    x = X1,
    y = X2,
    color = refinebio_specimen_part
  )
) +
  geom_point() # Plot individual points to make a scatterplot
```


## Complete differential expression analysis
```{r}
# Set seed for jitter plot
set.seed(999)

all.equal(colnames(final_mapping_df), metadata$refinebio_accession_code)

# Minimum counts cutoff
# Dataset is much bigger than the tutorial's dataset; ~80 times bigger
# Thus, we set the minimum count to 500 instead of 10 (~50 times more)
filtered_expression_df <- final_mapping_df %>%
  dplyr::filter(rowSums(.) >= 500)

# Dataset reduced to 18997 genes
nrow(filtered_expression_df)

# Change all counts to integers
gene_matrix <- round(filtered_expression_df)

deseq_results <- results(deseq)

# Get shrunken log fold change estimates to decrease noise and preserve large differences between groups
deseq_results <- lfcShrink(
  deseq,
  coef = 2,
  res = deseq_results
)

# Table of DESeq results
deseq_df <- deseq_results %>%
  as.data.frame() %>%
  rownames_to_column("Gene") %>%
  dplyr::mutate(threshold = padj < 0.05) %>%
  dplyr::arrange(dplyr::desc(log2FoldChange))

saveRDS(deseq_df, file = "deseq_df.RDS")

# Differentially expressed genes
deg_df <- deseq_df %>%
  filter(abs(log2FoldChange) > 1 & padj < 0.05)

saveRDS(deg_df, file = "deg_df.RDS")

# Save and summarize findings
write_tsv(deseq_df, file.path("results/SRP040561_DESeq2_results.tsv"))
write_tsv(deg_df, file.path("results/SRP040561_DEG.tsv"))

```

## Volcano plot
```{r}
volcano_plot <- EnhancedVolcano(
  deseq_df,
  lab = deseq_df$Gene,
  x = "log2FoldChange",
  y = "padj",
  pCutoff = 0.01
)

png(filename = "results/SRP040561_volcano_plot.png")
volcano_plot
dev.off()
```

# Heatmap
```{r}
# List of differentially expressed genes
filtered_deseq_df <- deseq_df %>% 
  filter(abs(log2FoldChange) > 1 & padj < 0.05)

heatmap_filtered_deseq_df <- filtered_deseq_df[(filtered_deseq_df$baseMean > 10) & (abs(filtered_deseq_df$log2FoldChange) > 1.5),]

heatmap_filtered_deseq_df <- heatmap_filtered_deseq_df %>%
  inner_join(collapsed_mapping_df, by = c("Gene" = "ensembl_gene_id"))
mat <- counts(deseq, normalized = T)[heatmap_filtered_deseq_df$Gene,]

mat.z <- t(apply(mat, 1, scale))
colnames(mat.z) <- colnames(mat)

heatmap_filtered_deseq_df <- heatmap_filtered_deseq_df %>%
  tibble::column_to_rownames("Gene")

h <- Heatmap(mat.z, cluster_rows = TRUE, cluster_columns = TRUE, column_labels = colnames(mat.z), name = "Z-score", row_labels = heatmap_filtered_deseq_df[rownames(mat.z),]$all_symbols)

png(filename = "results/SRP040561_heatmap.png", res = 250, width = 1000, height = 2000)
print(h)
dev.off()
```

# Enrichment Analysis 


## topGO

Run topGO for gene set enrichment analysis using Gene Ontology (GO).
GO terms can be split into three ontologies: Biological processes (BP), molecular functions (MF), and cellular components (CC).

```{r warning=FALSE}
# Create annotation file from collapsed_mapping_df
go_annots <- collapsed_mapping_df %>%
    select(ensembl_gene_id, go_ids)

write.table(go_annots,
            file = "annotations.txt",
            sep = "\t",
            row.names = FALSE,
            col.names = FALSE,
            quote = FALSE)

# Initialize variables for topGO
geneIDtoGO <- readMappings(file = "annotations.txt")
geneUniverse <- names(geneIDtoGO)
genesOfInterest <- as.character(deg_df$Gene)
geneList <- factor(as.integer(geneUniverse %in% genesOfInterest))
names(geneList) <- geneUniverse

topGORes <- function(ont) {
  # Create topGO object
  GOdata <- new("topGOdata",
              description = "topGO",
              ontology = ont,
              allGenes = geneList,
              geneSel = deg_df$Gene,
              annot = annFUN.gene2GO,
              gene2GO = geneIDtoGO)
  
  # Run a weighted Fisher's exact test, taking GO hierarchy into account
  resultFisher <- runTest(GOdata,
                        algorithm = 'weight01',
                        statistic = 'fisher')
  allGO <- usedGO(GOdata)
  topGO_res_init <- GenTable(GOdata,
                    weightFisher = resultFisher,
                    orderBy = 'weightFisher',
                    topNodes = length(allGO))

  # Adjust p-values using Benjamini-Hochberg correction
  padj <- round(p.adjust(topGO_res_init$weightFisher, method = "BH"), digits = 4)
  topGO_res <- cbind(topGO_res_init, padj)
  topGO_res <- topGO_res[order(topGO_res$padj),]
  
  return(topGO_res)
} 

topGO_BP <- topGORes("BP")
topGO_MF <- topGORes("MF")
topGO_CC <- topGORes("CC")

topGO_results <- rbind(topGO_BP, topGO_MF, topGO_CC)

# Save and summarize findings
saveRDS(topGO_results, file = "topGO_results.RDS")
write_tsv(topGO_results, file.path("results/SRP040561_topGO_results.tsv"))
```


## clusterProfiler

Run clusterProfiler for gene set enrichment analysis using Gene Ontology (GO). 

```{r warning=FALSE}
# Prepare input for clusterProfiler
gene_list <- deg_df$log2FoldChange
names(gene_list) <- deg_df$Gene
gene_list <- na.omit(gene_list)
gene_list <- sort(gene_list, decreasing = TRUE)

# BP, MF, and CC ontologies are all included
enrichment <- enrichGO(
  gene=deg_df$Gene,
  OrgDb = "org.Mm.eg.db", 
  keyType = "ENSEMBL",
  ont ="ALL",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
)

# Create data frame of gene set enrichment results
clusterProfiler_df <- as.data.frame(enrichment@result)

# Save and summarize findings
saveRDS(clusterProfiler_df, file = "clusterProfiler_results.RDS")
write_tsv(clusterProfiler_df, file.path("results/SRP040561_clusterProfiler_results.tsv"))
```

## gProfiler2
```{r}
gostres <- gost(query = filtered_deseq_df$Gene, organism = "mmusculus", ordered_query = FALSE, multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, measure_underrepresentation = FALSE, user_threshold = 0.05, correction_method = "bonferroni", domain_scope = "annotated", custom_bg = NULL, numeric_ns = "", sources = "GO", as_short_link = FALSE, highlight = TRUE)

gprof <- gostplot(gostres, capped = TRUE, interactive = FALSE)

png(filename = "results/SRP040561_gProfiler2.png")
print(gprof)
dev.off()
```


